---
title: "Introduccion Risk of AIV in Cuba during the last ten years"
author: "Beatriz Delgqdo Hernandez"
output: pdf_document
editor_options: 
  chunk_output_type: console
  markdown:
    references: "references"
      location: document
    prefix: "mydoc"
---
# Introduction {#intro}

## Introduction {#intro-intro}
The last decade has been characterized by changes in the population structure of different ecosystems due to the impact of different factors, including climatic crises and human-related actions (CITA). 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

MATERIALS AND METHODS


```{r library}
library(auk)
library(rgbif)
library(dplyr)
library(sf)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(readr)
sf_use_s2(FALSE)
```
DESCRIPTION OF STUDY AREA
```{r cuba layer}
map_proj <- st_crs(4326)

cuba_nac <- st_read("gadm41_CUB_0.shp") %>% 
  st_simplify(preserveTopology = FALSE, dTolerance = 0.01) %>% 
  st_transform(crs = map_proj) 
cuba_prov <- st_read("gadm41_CUB_1.shp") %>% 
  st_simplify(preserveTopology = FALSE, dTolerance = 0.01) %>% 
  st_transform(crs = map_proj) 
cuba_mun <- st_read("gadm41_CUB_2.shp") %>% 
  st_simplify(preserveTopology = FALSE, dTolerance = 0.01) %>% 
  st_transform(crs = map_proj) 
```
###DATA DESCRIPTION
**Migratory birds data collection**
Waterfowl data was extracted from eBird [eBird Basic Dataset (EBD)](https://ebird.org/science/download-ebird-data-products) website during 2013-2023 period. The data manipulation followed the procedure of [@Strimas-Mackey2023]. the bird counting was divided by migratory season per year


```{r function time}
# resolve namespace conflicts
select <- dplyr::select
# function to convert time observation to hours since midnight
time_to_decimal <- function(x) {
  x <- hms(x, quiet = TRUE)
  hour(x) + minute(x) / 60 + second(x) / 3600
}
```

####Extacting data from eBird
```{r ebird}
# # set ebd path
auk::auk_set_ebd_path("C:/abundance/BD/eBird_data/", overwrite = TRUE)
# output files
data_dir <- "data"
if (!dir.exists(data_dir)) {
  dir.create(data_dir)
}
# setup data directory
dir.create("data", showWarnings = FALSE)


# auk_set_ebd_path(path="C:/abundance/BD/eBird_data", overwrite = TRUE)
```


```{r loadingdata}

ebd_abahamensis <- auk_ebd(c("ebd_CU_whcpin_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_whcpin_smp_relAug-2023_sampling.txt"))
ebd_abahamensis <- ebd_abahamensis %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_dbicolor <- auk_ebd(c("ebd_CU_fuwduc_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_fuwduc_smp_relAug-2023_sampling.txt"))
ebd_dbicolor <- ebd_dbicolor %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_amewig <- auk_ebd(c("ebd_CU_amewig_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_amewig_smp_relAug-2023_sampling.txt"))
ebd_amewig <- ebd_amewig %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_bbwduc <- auk_ebd(c("ebd_CU_bbwduc_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_bbwduc_smp_relAug-2023_sampling.txt"))
ebd_bbwduc <- ebd_bbwduc %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_buffle <- auk_ebd(c("ebd_CU_buffle_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_buffle_smp_relAug-2023_sampling.txt"))
ebd_buffle <- ebd_buffle %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_buwtea <- auk_ebd(c("ebd_CU_buwtea_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_buwtea_smp_relAug-2023_sampling.txt"))
ebd_buwtea <- ebd_buwtea %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_cintea <- auk_ebd(c("ebd_CU_cintea_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_cintea_smp_relAug-2023_sampling.txt"))
ebd_cintea <- ebd_cintea %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_comter <- auk_ebd(c("ebd_CU_comter_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_comter_smp_relAug-2023_sampling.txt"))
ebd_comter <- ebd_comter %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_gadwal <- auk_ebd(c("ebd_CU_gadwal_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_gadwal_smp_relAug-2023_sampling.txt"))
ebd_gadwal <- ebd_gadwal %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_gnwtea <- auk_ebd(c("ebd_CU_gnwtea_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_gnwtea_smp_relAug-2023_sampling.txt"))
ebd_gnwtea <- ebd_gnwtea %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_gwfgoo <- auk_ebd(c("ebd_CU_gwfgoo_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_gwfgoo_smp_relAug-2023_sampling.txt"))
ebd_gwfgoo <- ebd_gwfgoo %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_hergul <- auk_ebd(c("ebd_CU_hergul_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_hergul_smp_relAug-2023_sampling.txt"))
ebd_hergul <- ebd_hergul %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_hoomer <- auk_ebd(c("ebd_CU_hoomer_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_hoomer_smp_relAug-2023_sampling.txt"))
ebd_hoomer <- ebd_hoomer %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_lessca <- auk_ebd(c("ebd_CU_lessca_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_lessca_smp_relAug-2023_sampling.txt"))
checklists <- read_sampling( ebd_lessca$file_sampling)
observations <- read_ebd(ebd_lessca$file)
# filter the checklist data
checklists <-  checklists%>% 
  filter(all_species_reported,
         protocol_type %in% c("Stationary", "Traveling"))

# filter the observation data
observations <- observations %>% 
  filter(all_species_reported,
         protocol_type %in% c("Stationary", "Traveling"))
ids <- intersect(checklists$checklist_id, observations$checklist_id)
f_sampling_lessca <- filter(checklists, checklist_id %in% ids)
f_ebd_lessca <- filter(observations, checklist_id %in% ids)

ebd_masduc <- auk_ebd(c("ebd_CU_masduc_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_masduc_smp_relAug-2023_sampling.txt"))
ebd_masduc <- ebd_masduc %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_musduc <- auk_ebd(c("ebd_CU_musduc_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_musduc_smp_relAug-2023_sampling.txt"))
ebd_musduc <- ebd_musduc %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_norpin <- auk_ebd(c("ebd_CU_norpin_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_norpin_smp_relAug-2023_sampling.txt"))
checklists <- read_sampling( ebd_norpin$file_sampling)
observations <- read_ebd(ebd_norpin$file)
# filter the checklist data
checklists <-  checklists%>% 
  filter(all_species_reported,
         protocol_type %in% c("Stationary", "Traveling"))

# filter the observation data
observations <- observations %>% 
  filter(all_species_reported,
         protocol_type %in% c("Stationary", "Traveling"))
ids <- intersect(checklists$checklist_id, observations$checklist_id)
f_sampling_norpin <- filter(checklists, checklist_id %in% ids)
f_ebd_norpin<- filter(observations, checklist_id %in% ids)


ebd_norsho <- auk_ebd(c("ebd_CU_norsho_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_norsho_smp_relAug-2023_sampling.txt"))
ebd_norsho <- ebd_norsho %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_rebmer <- auk_ebd(c("ebd_CU_rebmer_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_rebmer_smp_relAug-2023_sampling.txt"))
ebd_rebmer <- ebd_rebmer %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_redhea <- auk_ebd(c("ebd_CU_redhea_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_redhea_smp_relAug-2023_sampling.txt"))
ebd_redhea <- ebd_redhea %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_ribgul <- auk_ebd(c("ebd_CU_ribgul_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_ribgul_smp_relAug-2023_sampling.txt"))
ebd_ribgul <- ebd_ribgul %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_rinduc <- auk_ebd(c("ebd_CU_rinduc_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_rinduc_smp_relAug-2023_sampling.txt"))
ebd_rinduc <- ebd_rinduc %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_rudduc <- auk_ebd(c("ebd_CU_rudduc_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_rudduc_smp_relAug-2023_sampling.txt"))
ebd_rudduc <- ebd_rudduc %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_snogoo <- auk_ebd(c("ebd_CU_snogoo_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_snogoo_smp_relAug-2023_sampling.txt"))
checklists <- read_sampling( ebd_snogoo$file_sampling)
observations <- read_ebd(ebd_snogoo$file)
# filter the checklist data
checklists <-  checklists%>% 
  filter(all_species_reported,
         protocol_type %in% c("Stationary", "Traveling"))

# filter the observation data
observations <- observations %>% 
  filter(all_species_reported,
         protocol_type %in% c("Stationary", "Traveling"))
ids <- intersect(checklists$checklist_id, observations$checklist_id)
f_sampling_snogoo <- filter(checklists, checklist_id %in% ids)
f_ebd_snogoo<- filter(observations, checklist_id %in% ids)


ebd_wiwduc1 <- auk_ebd(c("ebd_CU_wiwduc1_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_wiwduc1_smp_relAug-2023_sampling.txt"))
ebd_wiwduc1 <- ebd_wiwduc1 %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()

ebd_wooduc <- auk_ebd(c("ebd_CU_wooduc_smp_relAug-2023.txt"),
               file_sampling = c("ebd_CU_wooduc_smp_relAug-2023_sampling.txt"))
ebd_wooduc <- ebd_wooduc %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()
```

```{r ebird-extract-filter}
# output files
data_dir <- "data"
if (!dir.exists(data_dir)) {
  dir.create(data_dir)
}
f_ebd_abahamensis <- file.path(data_dir, "ebd_abahamensis.txt")
f_sampling_abahamensis <- file.path(data_dir, "ebd_checklists_abahamensis.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_abahamensis)) {
  auk_filter(ebd_abahamensis, file = f_ebd_abahamensis, file_sampling = f_sampling_abahamensis)
}

f_ebd_amewig <- file.path(data_dir, "ebd_amewig.txt")
f_sampling_amewig <- file.path(data_dir, "ebd_checklists_amewig.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_amewig)) {
  auk_filter(ebd_amewig, file = f_ebd_amewig, file_sampling = f_sampling_amewig)
}

f_ebd_bbwduc <- file.path(data_dir, "ebd_bbwduc.txt")
f_sampling_bbwduc <- file.path(data_dir, "ebd_checklists_bbwduc.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_bbwduc)) {
  auk_filter(ebd_bbwduc, file = f_ebd_bbwduc, file_sampling = f_sampling_bbwduc)
}

f_ebd_buffle <- file.path(data_dir, "ebd_buffle.txt")
f_sampling_buffle <- file.path(data_dir, "ebd_checklists_buffle.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_buffle)) {
  auk_filter(ebd_buffle, file = f_ebd_buffle, file_sampling = f_sampling_buffle)
}

f_ebd_buwtea <- file.path(data_dir, "ebd_buwtea.txt")
f_sampling_buwtea <- file.path(data_dir, "ebd_checklists_buwtea.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_buwtea)) {
  auk_filter(ebd_buwtea, file = f_ebd_buwtea, file_sampling = f_sampling_buwtea)
}

f_ebd_cintea <- file.path(data_dir, "ebd_cintea.txt")
f_sampling_cintea <- file.path(data_dir, "ebd_checklists_cintea.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_cintea)) {
  auk_filter(ebd_cintea, file = f_ebd_cintea, file_sampling = f_sampling_cintea)
}

f_ebd_comter <- file.path(data_dir, "ebd_comter.txt")
f_sampling_comter <- file.path(data_dir, "ebd_checklists_comter.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_comter)) {
  auk_filter(ebd_comter, file = f_ebd_comter, file_sampling = f_sampling_comter)
}

f_ebd_dbicolor <- file.path(data_dir, "ebd_dbicolor.txt")
f_sampling_dbicolor <- file.path(data_dir, "ebd_checklists_dbicolor.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_dbicolor)) {
  auk_filter(ebd_dbicolor, file = f_ebd_dbicolor, file_sampling = f_sampling_dbicolor)
}

f_ebd_gadwal <- file.path(data_dir, "ebd_gadwal.txt")
f_sampling_gadwal <- file.path(data_dir, "ebd_checklists_gadwal.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_gadwal)) {
  auk_filter(ebd_gadwal, file = f_ebd_gadwal, file_sampling = f_sampling_gadwal)
}

f_ebd_gnwtea <- file.path(data_dir, "ebd_gnwtea.txt")
f_sampling_gnwtea <- file.path(data_dir, "ebd_checklists_gnwtea.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_gnwtea)) {
  auk_filter(ebd_gnwtea, file = f_ebd_gnwtea, file_sampling = f_sampling_gnwtea)
}

f_ebd_gwfgoo <- file.path(data_dir, "ebd_gwfgoo.txt")
f_sampling_gwfgoo <- file.path(data_dir, "ebd_checklists_gwfgoo.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_gwfgoo)) {
  auk_filter(ebd_gwfgoo, file = f_ebd_gwfgoo, file_sampling = f_sampling_gwfgoo)
}

f_ebd_hergul <- file.path(data_dir, "ebd_hergul.txt")
f_sampling_hergul <- file.path(data_dir, "ebd_checklists_hergul.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_hergul)) {
  auk_filter(ebd_hergul, file = f_ebd_hergul, file_sampling = f_sampling_hergul)
}

f_ebd_hoomer <- file.path(data_dir, "ebd_hoomer.txt")
f_sampling_hoomer <- file.path(data_dir, "ebd_checklists_hoomer.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_hoomer)) {
  auk_filter(ebd_hoomer, file = f_ebd_hoomer, file_sampling = f_sampling_hoomer)
}

# f_ebd_lessca <- file.path(data_dir, "ebd_lessca.txt")
# f_sampling_lessca <- file.path(data_dir, "ebd_checklists_lessca.txt")
# # only run if the files don't already exist
# if (!file.exists(f_ebd_lessca)) {
#   auk_filter(ebd_lessca, file = f_ebd_lessca, file_sampling = f_sampling_lessca)
# }

f_ebd_masduc <- file.path(data_dir, "ebd_masduc.txt")
f_sampling_masduc <- file.path(data_dir, "ebd_checklists_masduc.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_masduc)) {
  auk_filter(ebd_masduc, file = f_ebd_masduc, file_sampling = f_sampling_masduc)
}

f_ebd_musduc <- file.path(data_dir, "ebd_musduc.txt")
f_sampling_musduc <- file.path(data_dir, "ebd_checklists_musduc.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_musduc)) {
  auk_filter(ebd_musduc, file = f_ebd_musduc, file_sampling = f_sampling_musduc)
}

# f_ebd_norpin <- file.path(data_dir, "ebd_norpin.txt")
# f_sampling_norpin <- file.path(data_dir, "ebd_checklists_norpin.txt")
# # only run if the files don't already exist
# if (!file.exists(f_ebd_norpin)) {
#   auk_filter(ebd_norpin, file = f_ebd_norpin, file_sampling = f_sampling_norpin)
# }

f_ebd_norsho <- file.path(data_dir, "ebd_norsho.txt")
f_sampling_norsho <- file.path(data_dir, "ebd_checklists_norsho.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_norsho)) {
  auk_filter(ebd_norsho, file = f_ebd_norsho, file_sampling = f_sampling_norsho)
}

f_ebd_rebmer <- file.path(data_dir, "ebd_rebmer.txt")
f_sampling_rebmer <- file.path(data_dir, "ebd_checklists_rebmer.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_rebmer)) {
  auk_filter(ebd_rebmer, file = f_ebd_rebmer, file_sampling = f_sampling_rebmer)
}

f_ebd_redhea <- file.path(data_dir, "ebd_redhea.txt")
f_sampling_redhea <- file.path(data_dir, "ebd_checklists_redhea.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_redhea)) {
  auk_filter(ebd_redhea, file = f_ebd_redhea, file_sampling = f_sampling_redhea)
}

f_ebd_ribgul <- file.path(data_dir, "ebd_ribgul.txt")
f_sampling_ribgul <- file.path(data_dir, "ebd_checklists_ribgul.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_ribgul)) {
  auk_filter(ebd_ribgul, file = f_ebd_ribgul, file_sampling = f_sampling_ribgul)
}

f_ebd_rinduc <- file.path(data_dir, "ebd_rinduc.txt")
f_sampling_rinduc <- file.path(data_dir, "ebd_checklists_rinduc.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_rinduc)) {
  auk_filter(ebd_rinduc, file = f_ebd_rinduc, file_sampling = f_sampling_rinduc)
}

f_ebd_rudduc <- file.path(data_dir, "ebd_rudduc.txt")
f_sampling_rudduc <- file.path(data_dir, "ebd_checklists_rudduc.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_rudduc)) {
  auk_filter(ebd_rudduc, file = f_ebd_rudduc, file_sampling = f_sampling_rudduc)
}

# f_ebd_snogoo <- file.path(data_dir, "ebd_snogoo.txt")
# f_sampling_snogoo <- file.path(data_dir, "ebd_checklists_snogoo.txt")
# # only run if the files don't already exist
# if (!file.exists(f_ebd_snogoo)) {
#   auk_filter(ebd_snogoo, file = f_ebd_snogoo, file_sampling = f_sampling_snogoo)
# }

f_ebd_wiwduc1 <- file.path(data_dir, "ebd_wiwduc1.txt")
f_sampling_wiwduc1 <- file.path(data_dir, "ebd_checklists_wiwduc1.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_wiwduc1)) {
  auk_filter(ebd_wiwduc1, file = f_ebd_wiwduc1, file_sampling = f_sampling_wiwduc1)
}

f_ebd_wooduc <- file.path(data_dir, "ebd_wooduc.txt")
f_sampling_wooduc <- file.path(data_dir, "ebd_checklists_wooduc.txt")
# only run if the files don't already exist
if (!file.exists(f_ebd_wooduc)) {
  auk_filter(ebd_wooduc, file = f_ebd_wooduc, file_sampling = f_sampling_wooduc)
}


```


```{r ebird-zf}
ebd_zf_ebd_abahamensis <- auk_zerofill(f_ebd_abahamensis, f_sampling_abahamensis, collapse = TRUE)
ebd_zf_ebd_amewig <- auk_zerofill(f_ebd_amewig, f_sampling_amewig, collapse = TRUE)
ebd_zf_ebd_bbwduc <- auk_zerofill(f_ebd_bbwduc, f_sampling_bbwduc, collapse = TRUE)
ebd_zf_ebd_buffle <- auk_zerofill(f_ebd_buffle, f_sampling_buffle, collapse = TRUE)
ebd_zf_ebd_buwtea <- auk_zerofill(f_ebd_buwtea, f_sampling_buwtea, collapse = TRUE)
ebd_zf_ebd_cintea <- auk_zerofill(f_ebd_cintea, f_sampling_cintea, collapse = TRUE)
ebd_zf_ebd_comter <- auk_zerofill(f_ebd_comter, f_sampling_comter, collapse = TRUE)
ebd_zf_ebd_gadwal <- auk_zerofill(f_ebd_gadwal, f_sampling_gadwal, collapse = TRUE)
ebd_zf_ebd_dbicolor <- auk_zerofill(f_ebd_dbicolor, f_sampling_dbicolor, collapse = TRUE)
ebd_zf_ebd_gnwtea <- auk_zerofill(f_ebd_gnwtea, f_sampling_gnwtea, collapse = TRUE)
ebd_zf_ebd_gwfgoo <- auk_zerofill(f_ebd_gwfgoo, f_sampling_gwfgoo, collapse = TRUE)
ebd_zf_ebd_hergul <- auk_zerofill(f_ebd_hergul, f_sampling_hergul, collapse = TRUE)
ebd_zf_ebd_hoomer <- auk_zerofill(f_ebd_hoomer, f_sampling_hoomer, collapse = TRUE)
ebd_zf_ebd_lessca <- auk_zerofill(f_ebd_lessca, f_sampling_lessca, collapse = TRUE)#zero match
ebd_zf_ebd_masduc <- auk_zerofill(f_ebd_masduc, f_sampling_masduc, collapse = TRUE)
ebd_zf_ebd_musduc <- auk_zerofill(f_ebd_musduc, f_sampling_musduc, collapse = TRUE)
ebd_zf_ebd_norpin <- auk_zerofill(f_ebd_norpin, f_sampling_norpin, collapse = TRUE)#zero match
ebd_zf_ebd_norsho <- auk_zerofill(f_ebd_norsho, f_sampling_norsho, collapse = TRUE)
ebd_zf_ebd_rebmer <- auk_zerofill(f_ebd_rebmer, f_sampling_rebmer, collapse = TRUE)
ebd_zf_ebd_redhea <- auk_zerofill(f_ebd_redhea, f_sampling_redhea, collapse = TRUE)
ebd_zf_ebd_ribgul <- auk_zerofill(f_ebd_ribgul, f_sampling_ribgul, collapse = TRUE)
ebd_zf_ebd_rinduc <- auk_zerofill(f_ebd_rinduc, f_sampling_rinduc, collapse = TRUE)
ebd_zf_ebd_rudduc <- auk_zerofill(f_ebd_rudduc, f_sampling_rudduc, collapse = TRUE)
ebd_zf_ebd_snogoo <- auk_zerofill(f_ebd_snogoo, f_sampling_snogoo, collapse = TRUE)#zero match
ebd_zf_ebd_wiwduc1 <- auk_zerofill(f_ebd_wiwduc1, f_sampling_wiwduc1, collapse = TRUE)
ebd_zf_ebd_wooduc <- auk_zerofill(f_ebd_wooduc, f_sampling_wooduc, collapse = TRUE)

ebd_zf <- bind_rows(ebd_zf_ebd_abahamensis, ebd_zf_ebd_amewig, ebd_zf_ebd_bbwduc, ebd_zf_ebd_buffle, ebd_zf_ebd_buwtea, 
          ebd_zf_ebd_cintea, ebd_zf_ebd_comter, ebd_zf_ebd_gadwal, ebd_zf_ebd_dbicolor, ebd_zf_ebd_gnwtea,
          ebd_zf_ebd_gwfgoo, ebd_zf_ebd_hergul, ebd_zf_ebd_hoomer, ebd_zf_ebd_lessca, ebd_zf_ebd_masduc,
          ebd_zf_ebd_musduc, ebd_zf_ebd_norpin, ebd_zf_ebd_norsho, ebd_zf_ebd_rebmer, ebd_zf_ebd_redhea,
          ebd_zf_ebd_ribgul, ebd_zf_ebd_rinduc, ebd_zf_ebd_rudduc, ebd_zf_ebd_snogoo, ebd_zf_ebd_wiwduc1, 
          ebd_zf_ebd_wooduc)

```


```{r}
# clean up variables
ebd_zf <- ebd_zf %>% 
  mutate(
    # convert X to NA
    observation_count = if_else(observation_count == "X", 
                                NA_character_, observation_count),
    observation_count = as.integer(observation_count),
    # effort_distance_km to 0 for non-travelling counts
    effort_distance_km = if_else(protocol_type != "Traveling", 
                                 0, effort_distance_km),
    # convert time to decimal hours since midnight
    time_observations_started = time_to_decimal(time_observations_started),
    # split date into year and day of year
    year = year(observation_date),
    day_of_year = yday(observation_date)
  )

# Accounting for variation in detectability
# additional filtering
ebd_zf_filtered <- ebd_zf %>% 
  filter(
    # effort filters
    duration_minutes <= 6 * 60,
    effort_distance_km <= 10,
    # last 10 years of data
    year >= 2013,
    # 10 or fewer observers
     number_observers <= 10
    )%>% mutate(id=paste0(scientific_name, observation_date, longitude, latitude))
#Searching duplicated
# unique(ebd_zf_filtered %>% mutate(id=paste0(scientific_name, observation_date, longitude, latitude)))
```

```{r}
ebird <- ebd_zf_filtered %>%
  # tidyr::separate("observation_date", sep="-", into = c("year", "month", "day")) %>% 
  select(checklist_id, observer_id, sampling_event_identifier, id,
         scientific_name, iba_code,species_observed,
         observation_count, state_code, locality_id, locality, latitude, longitude,
         protocol_type, all_species_reported,
         observation_date, year, day_of_year, observation_date,
         time_observations_started, 
         duration_minutes, effort_distance_km,
         number_observers) %>% mutate(y=latitude, x=longitude) %>% 
  st_as_sf(coords = c("x", "y"), crs=st_crs(cuba_mun))

ebird<- st_intersection(ebird, cuba_mun %>% dplyr::select(GID_2), prepared = TRUE)



plot(st_geometry(cuba_nac))
plot(st_geometry(ebird), add=TRUE)


```

```{r finaldataseteBird}

write_csv(ebird, "ebird_data.csv", na = "")

```


#Extacting data from Gbif
```{r gbif}
gbif_an_ch <- read.delim("C:/Users/betty/BD/Gbif/0004563-230918134249559/occurrence.txt", header = TRUE, sep = "\t",
                  fill = TRUE, stringsAsFactors = TRUE) 
# head(occ)
gbif_an_ch <- gbif_an_ch[!is.na(gbif_an_ch$decimalLatitude) & !is.na(gbif_an_ch$decimalLongitude),]

gbif_an_ch <- gbif_an_ch %>% select(
  gbifID, recordedBy, recordedByID, occurrenceID, identificationID, level1Gid, level2Gid, level2Name, occurrenceStatus,
  eventDate, eventTime, year, month, day, dynamicProperties, locality, municipality, 
  phylum, kingdom, class, order, family, scientificName, individualCount, lifeStage, sex, habitat, decimalLatitude, 
  decimalLongitude, verbatimCoordinateSystem, coordinatePrecision
  ) %>% mutate(
    # convert NA to 0, assuming zero like absence of data
    individualCount = if_else(is.na(individualCount), 
                                0, individualCount),
    individualCount = as.integer(individualCount),
    observation_date=lubridate::make_date(year, month, day))

gbif_an_ch <- gbif_an_ch %>% rename(GID_1="level1Gid", GID_2="level2Gid") %>% filter(year>=2013) %>%  
  mutate(id = paste0(observation_date, scientificName, decimalLongitude,decimalLatitude))

gbif_data <- gbif_an_ch %>%  distinct(id, .keep_all = TRUE)
```
# CLEAN THE DATASET! ----
# mind that data often contain errors, so careful inspection and cleaning are necessary! 
# here we'll first remove records of absence or zero-abundance (if any):
```{r}
names(gbif_data)
sort(unique(gbif_data$individualCount))  # notice if some points correspond to zero abundance
sort(unique(gbif_data$occurrenceStatus))  # check for different indications of "absent", which could be in different languages! and remember that R is case-sensitive

write.csv(gbif_data, "gbif_data.csv")
```




